---
title: "Temporal Analysis of Danish Resistance Casualties, 1940–1945"
author: "Axel Landerholm and Marius Hatt"
date: 'Created on 05 May 2025 and updated `r format(Sys.time(), "%d %B, %Y")`'
output:  
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE)
```

```{r}
dir.create("figures")
```


## Load the resistance data

```{r}
library(tidyverse)
library(lubridate)
library(here)

# Read data
resistance <- read.csv2("data/resistance.csv")

head(resistance)
```

## Mutate data
```{r}
# Convert death column to three seperate columns
resistance <- resistance %>%
  mutate(date_of_death = dmy(date_of_death),   
         death_day = day(date_of_death),
         death_month = month(date_of_death),
         death_year = year(date_of_death))
```

```{r}
# Create a month-date (first day of the month)
resistance <- resistance %>%
  mutate(date_of_death = floor_date(date_of_death, "month"))

# Group by month
monthly_deaths <- resistance %>%
  group_by(date_of_death) %>%
  summarise(n = n()) %>%
  arrange(date_of_death)

# Show deadliest months
monthly_deaths %>%  
  arrange(desc(n))
```

## Create visual
```{r}
# Plot the months
graph1 <- ggplot(monthly_deaths, aes(x = date_of_death, y = n)) +
  geom_col(fill = "royalblue") +
  labs(title = "Resistance Deaths per Month During the Occupation",
    x = "Month",
    y = "Number of Deaths") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 15, face = "bold", hjust = 0.4)) +
    geom_vline(xintercept = as.Date("1940-04-09"), linetype = "dashed", color = "red") +
  annotate("text", x = as.Date("1940-04-09"), y = max(monthly_deaths$n, na.rm = TRUE),
           label = "Beginning of Occupation", 
           angle = 90, vjust = 2, hjust = 1.5, size = 3) +
  geom_vline(xintercept = as.Date("1943-08-28"), linetype = "dashed", color = "red") +
  annotate("text", x = as.Date("1943-08-28"), y = max(monthly_deaths$n, na.rm = TRUE),
           label = "End of Cooperation Policy", 
           angle = 90, vjust = 2, hjust = 1.5, size = 3) +
  geom_vline(xintercept = as.Date("1945-05-05"), linetype = "dashed", color = "red") +
  annotate("text", x = as.Date("1945-05-05"), y = max(monthly_deaths$n, na.rm = TRUE),
           label = "End of Occupation",
           angle = 90, vjust = 2, hjust = 1.8, size = 3)

# Visualize
graph1

# Save it
ggsave(plot = graph1, 
       here("figures", "death_per_month.png"), 
       height = 4, 
       width = 8,
       bg = "white",
       dpi = 300)
```

To add dotted lines for specific dates on the graph we put our code into ChatGPT and promted it with the following: "hvordan kan jeg markere annoteringer direkte på grafen?". We also asked ChatGPT to add code that made the graph cleaner looking.

## Region visual
```{r}
# Isolating regions 
region_counts <- resistance %>%
  mutate(region = ifelse(is.na(region), "(Missing Region Data)", region)) %>%
  count(region)
```

```{r}
# Plot the regions
graph2 <- ggplot(region_counts, aes(x = fct_reorder(region, n), y = n)) +
  geom_col(fill = "firebrick") +
  labs(title = "Resistance Deaths per Region",
       x = "Region",
       y = "Number of Deaths") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5))

# Show it 
graph2

# Save it
ggsave(plot = graph2,
       here("figures", "death_per_region.png"), 
       height = 4, 
       width = 8,
       bg = "white",
       dpi = 300)
```

We added our code for the graph to ChatGPT and promted it with the following: "lav grafen pænere og gør labelerne på x aksen mere læsbare". The result was that it used fct_reorder to organize the graph, as well as the code under theme().

## Resistance types by region visual
```{r}
# Find top 5 types of resistance
top5_types <- resistance %>%
  count(type_of_resistance, sort = TRUE) %>%
  slice_head(n = 5) %>%
  pull(type_of_resistance)

# Replace NA in region with "(Missing Region Data)"
resistance2 <- resistance %>%
  filter(type_of_resistance %in% top5_types) %>%
  mutate(region = ifelse(is.na(region), "(Missing Region Data)", region))

# Plot with the updated region
graph3 <- resistance2 %>%
  count(region, type_of_resistance) %>%
  ggplot(aes(x = type_of_resistance, y = n, fill = type_of_resistance)) +
  geom_col() +
  facet_wrap(~ region, ncol = 3) +
  labs(title = "Death Toll of Danish Resistance Fighters by Types of Resistance and Region",
       x = NULL,
       y = "Number of Deaths",
       fill = "Type of Resistance") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.1),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank())

# Show it
graph3

# Save it
ggsave(plot = graph3,
       here("figures", "death_region_type.png"), 
       height = 5, 
       width = 8,
       bg = "white",
       dpi = 300)

```

We  added our code for the previous graph in ChatGPT and promted it with the following: "hvordan laver jeg et facetplot i rstudio, hvor den er delt op mellem de 7 regioner i forhold til type_of_resistance"
